---
title: "Context Factorisation with tensor-cell2cell"
author:
  - name: Daniel Dimitrov
    affiliation:
        - Saezlab, Heidelberg University
    email: daniel.dimitrov@uni-heidelberg.de
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
      self_contained: true
      toc: true
      toc_float: true
      toc_depth: 3
      code_folding: show
vignette: >
  %\VignetteIndexEntry{Contex Factorization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Tensor decomposition as proposed in the [tensor_cell2cell](https://www.nature.com/articles/s41467-022-31369-2#ref-CR5) paper, 
enables us to decipher context-driven intercellular communication by simultaneously
accounting for an unlimited number of "contexts". These contexts could represent
samples coming from longtidinal sampling points, multiple conditions, or cellular niches.

The power of tensor-cell2cell is in its ability to decompose latent patterns of
intercellular communication in an untargeted manner, in theory being able to handle
cell-cell communication results coming from any experimental design, regardless
of its complexity.

Simply put, tensor_cell2cell uses LIANA's output `by sample` to build a 4D tensor,
represented by 1) contexts, 2) interactions, 3) sender, and 4) receiver cell types.
This tensor is then decomposed into a set of factors, which can be interpreted as
low dimensionality latent variables (vectors) that capture the
CCC patterns across contexts.
  

Here, we will combine `LIANA` with `tensor_cell2cell` to decipher potential ligand-receptor
interaction changes.
As a simple example, we will look at ~14000 PBMCs from 8 donors,
each before and after IFN-beta stimulation (GSE96583; obtained via [ExperimentHub](https://bioconductor.org/packages/release/bioc/html/ExperimentHub.html) & [muscData](https://bioconductor.org/packages/release/data/experiment/html/muscData.html)).
Note that by focusing on PBMCs, for the purpose of this tutorial, we assume that
coordinated events occur among them.


This tutorial was heavily influenced by the [tutorials of tensor_cell2cell](https://earmingol.github.io/cell2cell/tutorials/ASD/01-Tensor-Factorization-ASD/).

Any usage of `liana` x `tensor_cell2cell` should logically cite both articles,
and in particular `tensor_cell2cell` (see reference at the bottom).

Load required libraries
```{r, message=FALSE, warning=FALSE}
library(tidyverse, quietly = TRUE)
library(SingleCellExperiment, quietly = TRUE)
library(reticulate, quietly = TRUE)
library(magrittr, quietly = TRUE)
library(liana, quietly = TRUE)
library(ExperimentHub, quietly = TRUE)
```

### Request Data and Preprocess

##### Request Data
```{r, message=FALSE, warning=FALSE}
eh <- ExperimentHub()
# Get Data
(sce <- eh[["EH2259"]])
```


##### Preprocess
```{r, message=FALSE, warning=FALSE}
# basic feature filtering
sce <- sce[rowSums(counts(sce) >= 1) >= 5, ]

# basic outlier filtering
qc <- scater::perCellQCMetrics(sce)

# remove cells with few or many detected genes
ol <- scater::isOutlier(metric = qc$detected, nmads = 2, log = TRUE)
sce <- sce[, !ol]

# Remove doublets
sce <- sce[, sce$multiplets=="singlet"]

# Set rownames to symbols
rownames(sce) <- rowData(sce)$SYMBOL

# log-transform
sce <- scuttle::logNormCounts(sce)

# Create a label unique for every sample
sce$context <- paste(sce$stim, sce$ind, sep="|")
```


##### Ensure Consitency across Cell identities

To obtain consistent CCC patterns across samples, we need to make sure that
the cell identities are stable. Namely,

```{r, message=FALSE, warning=FALSE, fig.width=16, fig.height=12}
# Plot
sce %>%
  get_abundance_summary(sample_col = "context",
                        idents_col = "cell", 
                        min_cells = 10, # min cells per sample
                        min_samples = 3, # min samples
                        min_prop = 0.2 # min prop of samples
                        ) %>%
  plot_abundance_summary()

# filter non abundant celltypes
sce <- liana::filter_nonabundant_celltypes(sce,
                                           sample_col = "context",
                                           idents_col = "cell")
```


### Run liana for on each individual sample.

In order to construct the tensor, we first need to obtain CCC predictions for 
each sample. In this case, we use SingleCellSignalR scores, as they are
regularized, and theory directly comparable between any dataset. One can use any
method in LIANA /w tensor as they were previously shown to yield consistent
results (Armingol & Baghdassarian, 2022).

Note that `liana_bysample` works with `SingleCellExperiment` alone, if you 
wish to use Seurat, please use the `as.SingleCellExperiment` function.
```{r runliana, eval=TRUE, message=FALSE, warning=FALSE}
# Run LIANA by sample
sce <- liana_bysample(sce = sce,
                      sample_col = "context",
                      idents_col = "cell",
                      method = "sca", # we use SingleCellSignalR's score alone
                      expr_prop = 0, # expression proportion threshold
                      inplace=TRUE # saves to sce
                      )
```



We expect to see a successful LIANA run for each sample/context.
```{r}
summary(sce@metadata$liana_res)
```


## Cell-cell Communication Tensor Decomposition
Here, we call tensor_cell2cell.

This function will first format the ligand-receptor scores per sample into a 
4 Dimensional tensor.

It will then estimate the number of factors to which the tensor
will be decomposed (set rank to `NULL`, for the sake of computational speed,
I have pre-calculated the `rank` and I explicitly set it to 7 here). 
Optimal rank estimation can be computationally demanding, thus in this case
the number ranks was pre-determined.

One can think of this as a higher-order non-negative matrix factorization where
the factors can be used to reconstruct the tensor. We refer the user to the
publication of [tensor-cell2cell](https://www.nature.com/articles/s41467-022-31369-2) for further information.

Note that by default LIANA will set up a conda environment with `basilisk`
(if conda_env is NULL), the user can alternatively specify the name of a `conda_env`
with cell2cell to be called via reticulate.

```{r, warning=FALSE}
sce <- liana_tensor_c2c(sce = sce,
                        score_col = 'LRscore',
                        rank = 7,  # set to None to estimate!
                        how='outer'  #  defines how the tensor is built
                        )
```


The `how` parameter plays is essential in how we treat missing cell types and
interactions. In this scenario, we use `outer`, which will only decompose CCC
across *any* cell identities and interactions in the dataset. Alternatively,
one could change it to 'inner' then CCC paterns will be decomposed patterns
across cell identities & interactions shared between *all* samples/contexts. 

`cell2cell_tensor` also accepts additional parameters to further tune 
tensor decomposition, but these are out of scope for this tutorial.
Stay tuned for a set of tutorials with comprehensive application of liana with tensor!

### Results Overview

Upon a seccusseful run cell2cell_tensor will decompose the tensor into a set of factors, 
or four vectors corresponding to the initial dimensions of the tensor.

* `contexts` - the factor scores assigned to each sample
* `interactions` - the interaction loadings per factor
* `senders` - loadings for sender/source cell identities
* `receivers` - loadings for receivers/target cell identities
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# get the factors
factors <- get_c2c_factors(sce)

# show them
glimpse(factors)
```


Here, we examine the behavior of the different dimensions across the factors.
When we look at the contexts (samples) loadings `Factor.4` seems to be notably
different between the groups. We also see that the loadings for the "Sender" and
"Receiver" cells have a relatively uniform distribution, suggesting that most
cell types are involved in the CCC events that distinguish the conditions.

```{r, message=FALSE, warning=FALSE, fig.width=16, fig.height=12}
# Plot overview
plot_c2c_overview(sce, group_col="stim")
```

### Statistical comparison of Communication Patterns
Here, we see that the communication patterns (or context loadings) identified 
statistically significant patterns before and after stimulation.

These factors thus represent differences in the ligand-receptor interactions
as well as cell types participating in cell-cell communication before
and after IFN-beta stimulation.


```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# Get all boxplots
all_facts_boxes <- plot_context_boxplot(sce,
                                        group_col = "stim",
                                        test="t.test", # applicable only to two groups
                                        paired=TRUE #! Is this the case for your data?
                                        )

# Combine all boxplots
require(patchwork)
wrap_plots(
    all_facts_boxes,
    ncol=4) +
    plot_layout(guides = "collect") & theme(legend.position = 'bottom') +
    theme(plot.tag = element_text(face = 'bold', size = 16)
          )

```



### Contexts Heatmap

Again, we see a clear seperation between STIM and CTRL, further suggesting the
relevance of the changes in the inferred interactions.
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}

plot_context_heat(sce, group_col = "stim")

```


## Cell-cell pairs with high potential of interaction

Here we use the product of the `source` and `target` loadings to generate the
heatmap of potential celltype-celltype pair relationships, which contribute to
`Factor.4`.
```{r ccpairs, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}

plot_c2c_cells(sce,
               factor_of_int = "Factor.4",
               name = "Loadings \nProduct")
```

### Gini Coefficients of Factor-specific Communicating Sender and Receiver cell types

Gini coefficients range from 0 to 1, and are measures of dispersion,
typically used to measure inequality.

Here, the Gini coefficient is used to measure the imbalance of communication 
in terms of the cell types.

Gini coefficient of 1 would suggests that there is a single cell type contributing
to the communication patterns within a factor, while a value of 0 suggest perfect
equality between the cell type.

When we focus on `Factor.4`, we see that the both source/sender and target/receiver
cell types have nearly uniform contributions.

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# Get loadings for source/sender Cell types
calculate_gini(factors$senders)
```


```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# Get loadings for target/receiver Cell types
calculate_gini(factors$receivers)
```

## LR loadings Heatmap
We can see the LRs involved across contexts.

Though, perhaps since in this case Factors 4 and 5 are those associated with
the stimulation, perhaps it's best to focus on those.
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
plot_lr_heatmap(sce,  n = 5, cluster_columns=FALSE)
```
  
  
## LR loadings Pathway Enrichment

Here, we will use pathways from [PROGENy](https://github.com/saezlab/progeny),
together with [decoupleR](https://github.com/saezlab/decoupleR), to do an 
enrichment analysis on the LR loadings.

`PROGENy` represent a data-driven signatures for 14 pathways, while `decoupleR`
is an framework with multiple enrichment methods.

Since PROGENy was originally intended to be used with genes,
we need to reformat it to match 
ligand-receptor predictions. Namely, `generate_lr_geneset` will assign a LR
to a specific pathway, only if both the ligand and receptor, as well as all
their subunits, are associated with the same pathway. And also in the case of
weighted gene sets (like progeny), all entities have the same weight sign.

```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}

# obtain progeny gene sets
progeny <- decoupleR::get_progeny(organism = 'human', top=5000) %>%
    select(-p_value)

# convert to LR sets
progeny_lr <- generate_lr_geneset(factors$interactions,
                                  resource = progeny)

progeny_lr
```


##### Enrichment Dotplot

Here, we see that LRs associated with the JAK-STAT gene in progeny are
enriched in `Factor.4`.
```{r, fig.width=12, fig.height=8, message=FALSE, warning=FALSE}
# interaction loadings to matrix
mat <- factors$interactions %>%
    column_to_rownames("lr") %>%
  as.matrix()

# run enrichment analysis with decoupler
# (we fit a univariate linear model for each gene set)
# We don't consider genesets with minsize < 10
res <- decoupleR::run_ulm(mat = mat,
                          network = progeny_lr,
                          .source = "set",
                          .target = "lr",
                          minsize=10) %>%
  mutate(p_adj = p.adjust(p_value, method = "fdr"))

res %>% # sig/isnig flag
  mutate(significant = if_else(p_adj <= 0.05, "signif.", "not")) %>%
  ggplot(aes(x=source, y=condition, shape=significant,
             colour=score, size=-log10(p_value+1e-36))) +
  geom_point() +
  scale_colour_gradient2(high = "red", low="blue") +
  scale_size_continuous(range = c(3, 12)) +
  scale_shape_manual(values=c(21, 16)) +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x="Pathway",
       y="Factor",
       colour="Activity"
       )

```

##### LRs driving enrichment of JAK-STAT in Factor.4

```{r, echo=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}
# Plot LRs associated with Estrogen
lrs <-  factors$interactions %>%
  left_join(progeny_lr, by="lr") %>%
  filter(set=="JAK-STAT") %>%
  select(lr, set, mor, loading = Factor.3) %>%
  mutate(lr = gsub(as.character(str_glue("\\^")), " -> ", lr)) %>%
  mutate(weight = if_else(mor >= 0, "positive", "negative"))
lrs %>%
  # only label those that are > x
  mutate(lr = if_else(loading>=0.001 & abs(mor) > 2, lr, "")) %>%
  ggplot(aes(x=mor, y=loading, colour=weight)) +
  # label only top 20
  stat_smooth(method = "lm", col = "red") +
  geom_point(alpha = 0.5) + 
  ggrepel::geom_label_repel(aes(label = lr)) +
  theme_bw(base_size = 15) +
  scale_colour_manual(values = c("royalblue3", "red")) +
  labs(x="Pathway Weight", y="LR Loading")
```



## Citation

To Cite tensor_cell2cell:
Armingol, E., Baghdassarian, H.M., Martino, C., Perez-Lopez, A., Aamodt, C., Knight, R. and Lewis, N.E., 2022. Context-aware deconvolution of cell–cell communication with Tensor-cell2cell. Nature Communications, 13(1), pp.1-15.

To Cite LIANA:
```{r liana citation, message = FALSE, echo=FALSE}
citation("liana")
```

## Session information
```{r session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```

