---
title: "LIANA++ for Benchmarks and Developers"
author: "Daniel Dimitrov"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{LIANA++ for Benchmarks and Developers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIANA ++

To facilitate further method development and the benchmarks of current and future methods,
in this vignette we provide further instructions relevant for method developers.


## Load prerequisites
```{r load_prereq, message=FALSE, warning=FALSE}
require(liana)
require(tibble)
require(purrr)
require(magrittr)

liana_path <- system.file(package = "liana")
testdata <-
    readRDS(file.path(liana_path , "testdata", "input", "testdata.rds"))
```


## `liana_pipe` and Contributions
To facilitate method development, we provide a guide how to make use of the architecture already implemented in LIANA.

First, we start by converting the object to SCE (if Seurat) and basic checks via `liana_rep`.
Then we generate a LR_res table with many relevant columns, such as stats for ligands and receptors per cell type.
Note that liana_pipe also takes complexes into account by dissociating them into subunits (see ?`liana:::decomplexify`).
These are later re-assembled back into complexes using the `recomplexify` function.
```{r lr_cpdb, message=FALSE, warning=FALSE}
sce <- liana_prep(testdata)

# Obtain LR summary res with CellPhoneDB and dissociate complexes
lr_res <- liana_pipe(sce,
                     op_resource = select_resource("CellPhoneDB")[[1]] %>%
                       liana:::decomplexify(),
                     base = 2 # here we assumed log2-transformation
                     # of the library-normalized counts
                     ) %>%
    glimpse()
```
``


The LR result summary is then passed to the `get_*` functions then provide ways
to summarize this information into a score (`LRscore` in this case) that can be
used to the prioritize interactions.

We also need to pass the function that we wish to use to account for heteromeric complexes
```{r lr_get, message=FALSE, warning=FALSE}
get_sca(lr_res, complex_policy="mean") %>% glimpse
```
`get_*` functions refer to `liana_call` and `liana_score` functions behind the scenes, so please refer to those for more info.



## Install LIANA++
Install a LIANA++ environment with all original software as a way to minimize bias for benchmark studies.

LIANA++ uses CCC methods from both R and Python, as such dependencies for both need to be installed.

To install the LIANA++ framework run the following code in R:
```{r install_r, eval=FALSE}
require(devtools)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")

Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = TRUE) # ignore warning from iTALK
BiocManager::install("ComplexHeatmap") # required for Connectome
devtools::install_github('saezlab/OmnipathR@ff3ad88e3915747e1b557bf44ac5396f9525dd7e') # install 4.0 version of OmnipathR

# install tools
devtools::install_github("sqjin/CellChat")
devtools::install_github('msraredon/Connectome', ref = 'master')
devtools::install_github("Coolgenome/iTALK", build_vignettes = TRUE)
# A modified version of SingleCellSignalR (SCA) that enables external resources
devtools::install_github(repo = "saezlab/SingleCellSignalR_v1",
                         subdir = "SingleCellSignalR")

# Finally, install LIANA
devtools::install_github('saezlab/liana')
```



Squidpy and NATMI are written in Python, as such a python environment with
the prerequisites of these methods needs to be set up.

If you also wish to run the CellPhoneDB re-implementation from Squidpy, please set up
a [conda](https://docs.conda.io/projects/conda/en/latest/user-guide/index.html)
environment by running the following lines in the terminal:
```{bash, eval = FALSE}
conda create -n liana_env
conda activate liana_env
conda install -c anaconda python=3.8.5
pip install squidpy
```


Please use the [.yml](https://github.com/saezlab/liana/blob/master/liana_env.yml) file to set up a conda environment:
```
conda env create -f liana_env.yml
```

Note! NATMI and Squidpy are set by default to look for this conda environment ("liana_env").

#### To use NATMI with LIANA, clone the modified NATMI repo into liana path by running the following in the terminal:
```{sh, eval=FALSE}
cd *insert fullpath*/liana
git clone https://github.com/saezlab/NATMI
```

##### Note that, NATMI was forked from [asrhou/NATMI](https://github.com/asrhou/NATMI.git) and changed to be agnostic of the working directory when loading the resources.


#### If needed, use the following to locate the liana package
```{r get_liana, eval=FALSE}
system.file(package = "liana")
# "/home/user/R/x86_64.../4.0/liana"
```
It might be easier to just also clone the liana repo.

#### `call_*` functions
Use the *call_* functions to run the pipelines of the original methods, rather than their re-implementations.
These functions work with Seurat objects and they are instead converted to SCE. These are largely getting **deprecated**
and will be moved to a seperate repository in the future.
```{r liana_call, message=FALSE, warning=FALSE, print=FALSE}
# RUN cellchat alone with OmniPath's Consensus
sca_res <- call_sca(
  testdata,
  op_resource = select_resource('Consensus')[[1]]
  )

# Show CellChat Results
sca_res
```

Similarly one could call all `call_*` functions via `liana_wrap`
```{r liana_calls, message=FALSE, warning=FALSE, print=FALSE}
liana_calls <- liana_wrap(testdata,
                          method = c('call_connectome',
                                     'call_sca',
                                     'call_italk',
                                     'call_natmi'
                                     ),
                          resource = c("Default")) # Run with Default Resources
liana_calls %>% liana_aggregate() # Connectome finds no significant hits here
# Note that for italk we use mean(ligand.logFC, receptor.logFC) as a way to compare
# to the other methods
```


## Session information
```{r session_info, echo=TRUE}
options(width = 120)
sessioninfo::session_info()
```

