---
title: "LIANA with Intracellular Omnipath"
author:
  - name: Daniel Dimitrov
    affiliation:
        - Saezlab, Heidelberg University
    email: daniel.dimitrov@uni-heidelberg.de
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
      self_contained: true
      toc: true
      toc_float: true
      toc_depth: 3
      code_folding: show
vignette: >
  %\VignetteIndexEntry{LIANA with Intracellular Omnipath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Making use of [Intracellular OmniPath](https://omnipathdb.org/#about)

This vignette showcases how to one could obtain and subsequently use the 
information from OmniPath's intracellular components with LIANA. 


## Load Prerequisites
```{r load_prereq, message=FALSE, warning=FALSE}
require(tidyverse)
require(liana)
require(purrr)
require(magrittr)

liana_path <- system.file(package = "liana")
testdata <-
    readRDS(file.path(liana_path , "testdata", "input", "testdata.rds"))
```

## Filter OmniPath and LIANA by specific functional annotations
   
Here, we will assume that we have strong reason to believe that the RTK pathway
is particularly relevant in our study and we are hence solely interested in
interactions from the `RKT` pathway. Thus, we will filter LIANA and OmniPath to 
only interactions associated with the `RKT` pathway from [SignaLink](https://link.springer.com/protocol/10.1007/978-1-4939-8618-7_3).
*Signalink* itself is a signaling pathway resource with multi-layered regulatory networks.

First, let's obtain SignaLink
```{r load_signalink}
signalink_pathways <-
    OmnipathR::import_omnipath_annotations(
        resources = 'SignaLink_pathway',
        entity_types = 'protein',
        wide = TRUE
    ) %>%
  select(genesymbol, pathway)
```

Then we obtain our resource of choice and run liana.
The resource can also be customized, to do so please refer to  
 [LIANA's Customize CCC OmniPath tutorial](https://saezlab.github.io/liana/articles/liana_custom_op.html) 

```{r liana_omni, message = FALSE, print = FALSE}
# Obtain resource and format it
liana_omni <- select_resource("OmniPath")[[1]] %>%
    liana:::decomplexify()

# Run liana with our resource
liana_res <- liana_wrap(testdata,
                        resource='custom',
                        external_resource = liana_omni) %>%
  # Note! here we work with the 'minimum' subunits
  liana_aggregate(join_cols = c("source", "target",
                                "ligand", "receptor")) # aggregate
```

### Filter LIANA output to only RTK interactions
Here, we assume that we only are only interested in from the RTK pathway.
```{r filter_by_liana}
liana_rtk <- liana_res %>% 
  left_join(signalink_pathways, by=c("ligand"="genesymbol")) %>%
  rename(ligand_pathway = pathway) %>% 
  left_join(signalink_pathways, by=c("receptor"="genesymbol")) %>%
  rename(receptor_pathway = pathway) %>%
  filter(ligand_pathway == "Receptor tyrosine kinase") %>%
  filter(receptor_pathway == "Receptor tyrosine kinase")

# Show only interactions in which BOTH the ligand and the receptor
# are associated with the RTK pathway
liana_rtk
```

### Run LIANA with only specific interactions
One could also do the reverse - i.e. filter the resource in regards to a given
functional annotation terms (e.g. JAK/STAT, WNT, and RTK), and then run LIANA
```{r filter_by_op}
pathways_of_interest <- c("JAK/STAT", "Receptor tyrosine kinase", "WNT")

# We join the functional annotations to OmniPath
# and retain ONLY interactions in which BOTH the ligand and receptor 
# are associated with RTK pathways
rkt_omni <- liana_omni %>% 
  left_join(signalink_pathways, by=c("source_genesymbol"="genesymbol")) %>%
  rename(source_geneset = pathway) %>%
  left_join(signalink_pathways, by=c("target_genesymbol"="genesymbol")) %>%
  rename(target_geneset = pathway) %>%
  filter(source_geneset %in% pathways_of_interest) %>%
  filter(target_geneset %in% pathways_of_interest)

# We can then again run LIANA with the RKT-associated interactions alone
liana_rtk_omni <- liana_wrap(testdata,
                             resource='custom',
                             external_resource = rkt_omni) %>%
  liana_aggregate(join_cols = c("source", "target",
                                "ligand", "receptor"))

liana_rtk_omni
```


## Perform an over-representation representation analysis (ORA) on predicted LR interactions 
Here, (a bit more advanced) we will attempt to obtain over-represented MSigDB  
gene sets in preferentially ranked ligand-receptor interactions from LIANA.

### First, format LIANA's output
```{r liana_wrap, message = FALSE, print = FALSE}
# we take liana from above and keep only the columns that we need
# in this case we will use the aggragate for all methods
liana_form <- liana_res %>%
  select(source, ligand, target,
         receptor, aggregate_rank) %>%
  # one could treat the aggragate rank as 
  # the probability to observe an interaction
  # as preferentially highly ranked
  # one could also p.adjust
  # mutate(aggregate_rank = p.adjust(aggregate_rank)) %>%
  filter(aggregate_rank <= 0.05)
liana_form

```

### Obtain MSigDB Genesets via OmniPathR - takes a while :)
Here we focus on the [Hallmarks gene set](http://www.gsea-msigdb.org/gsea/msigdb/collection_details.jsp#H)
```{r load_msig}
msigdb <- OmnipathR::import_omnipath_annotations(resources = "MSigDB",
                                                 wide = TRUE) %>%
  filter(collection=="hallmark") %>%
  select(genesymbol, geneset)
```

### Establish the Resource as the background/universe
```{r establish_universe}
# Here, we join pathways associated with ligand and receptors, separately
# Note that here we consider only matches to ligands and receptors from OmniPath as the background universe
omni_universe <- liana_omni %>% 
  select(source_genesymbol, target_genesymbol) %>%
  left_join(msigdb, by=c("source_genesymbol"="genesymbol")) %>%
  rename(source_geneset = geneset) %>%
  left_join(msigdb, by=c("target_genesymbol"="genesymbol")) %>%
  rename(target_geneset = geneset)

# Establish Ligand Universe
entity_total <- length(unique(omni_universe$source_genesymbol))
ligand_universe <- omni_universe %>%
  select(ligand = source_genesymbol, geneset = source_geneset) %>%
  na.omit() %>%
  distinct() %>%
  group_by(geneset) %>%
  mutate(geneset_n = n()) %>%
  ungroup() %>%
  arrange(desc(geneset_n)) %>%
  mutate(background = entity_total)


# Bind Distinct liana ligand hits to the universe
# Get Distinct Ligands in top hits
liana_ligands <- liana_form %>%
  select(source, ligand) %>%
  distinct() %>%
  mutate(distinct_hits = n())

# Perform hypergeometric test
liana_hyperenrich <- ligand_universe %>%
  left_join(liana_ligands, by=c("ligand")) %>%
  na.omit() %>%
  group_by(source, geneset) %>%
  # count ligands in geneset
  mutate(ligands_in_gs = n()) %>% # aka x/q
  rowwise() %>%
  mutate(pval = phyper(q = ligands_in_gs, 
                       m = geneset_n,
                       n = background-geneset_n,
                       k = distinct_hits, lower.tail = FALSE)) %>%
  ungroup() %>%
  arrange(pval)

# Note that the ligand coverage of some gene sets is simply too limited
# So, one could consider not limiting the results to the ligands alone
liana_hyperenrich
```
### Format Hypergeometric test results
```{r ora_res}
liana_hyperenrich %>%
  select(geneset, source, pval) %>%
  distinct() %>%
  mutate(adj_pval = p.adjust(pval, method = "fdr")) %>%
  arrange(adj_pval)
```

Done! Check whether any categories are over-represented (by celltype) in our dataset.

Kind reminder that this analysis is intended for representation purposes alone. 
We usually favour the use of the whole genome for any enrichment and pathway analyses,
and kindly refer to the user to the rest of [our tools](https://saezlab.org/?#tools), which can be highly complementary
when interpretting the ligand-receptor predictions from LIANA.


We also kindly refer the user to [OmniPathR](https://github.com/saezlab/OmnipathR), and the [OmniPath website](https://omnipathdb.org/#about) for
more information how to make full use of OmniPath!


## Session information
```{r session_info, echo=TRUE}
options(width = 120)
sessioninfo::session_info()
```
