---
title: "Context Factorisation with tensor-cell2cell"
author:
  - name: Daniel Dimitrov
    affiliation:
        - Saezlab, Heidelberg University
    email: daniel.dimitrov@uni-heidelberg.de
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
      self_contained: true
      toc: true
      toc_float: true
      toc_depth: 3
      code_folding: show
vignette: >
  %\VignetteIndexEntry{Contex Factorization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Tensor decomposition as proposed in the [tensor_cell2cell](https://www.nature.com/articles/s41467-022-31369-2#ref-CR5) paper, 
enables us to decipher context-driven intercellular communication by simultaneously
accounting for an unlimited number of "contexts". These contexts could represent
samples coming from longtidinal sampling points, multiple conditions, or cellular niches.

The power of tensor-cell2cell is in its ability to decompose latent patterns of
intercellular communication in an untargeted manner, in theory being able to handle
cell-cell communication results coming from any experimental design, irregardless
of its complexity.

Simply put, tensor_cell2cell uses LIANA's output `by sample` to build a 4D tensor,
represented by 1) contexts, 2) interactions, 3) sender, and 4) receiver cell types.
This tensor is then decomposed into a set of factors, which can be interpreted as
low dimensionality latent variables that capture the CCC patterns across contexts.
  

Here, we will combine `LIANA` with `tensor_cell2cell` to decipher potential ligand-receptor
interaction changes.
As a simple (possibly overoptimistic) example, we will look at ~14000 PBMCs from 8 donors,
each before and after IFN-beta stimulation (GSE96583; obtained from: https://scelvis.readthedocs.io/en/latest/tutorial_analysis.html).


This tutorial was heavily influenced by the [tutorials of tensor_cell2cell](https://earmingol.github.io/cell2cell/tutorials/ASD/01-Tensor-Factorization-ASD/).

Any usage of `liana` x `tensor_cell2cell` should logically cite both articles,
and in particular `tensor_cell2cell` (see reference at the bottom).

Load required libraries
```{r, message=FALSE, warning=FALSE}
library(tidyverse, quietly = TRUE)
library(SingleCellExperiment, quietly = TRUE)
library(reticulate, quietly = TRUE)
library(magrittr, quietly = TRUE)
library(liana, quietly = TRUE)
```

Load processed data.
```{r loadstuff, message=FALSE, warning=FALSE}
sce <- readRDS(url("https://zenodo.org/record/6826420/files/pbmc_sce.RDS?download=1"))

# library(ExperimentHub)
# eh <- ExperimentHub()
# # Get Data
# (sce <- eh[["EH2259"]])
```


### Run liana for on each individual sample.

In order to construct the tensor, we first need to obtain CCC predictions for 
each sample. In this case, we use SingleCellSignalR scores, as they are
regularized, and theory directly comparable between any dataset.

Note that `liana_bysample` works with `SingleCellExperiment` alone, if you 
wish to use Seurat, please use the `as.SingleCellExperiment` function.
```{r runliana, eval=TRUE, message=FALSE, warning=FALSE}
# Run LIANA by sample
sce <- liana_bysample(sce = sce,
                      sample_col = "Donor", 
                      condition_col = "stim",
                      idents_col = "cluster",
                      method = "sca",
                      expr_prop = 0) # here, we lower liana's filtering
```



```{r, include=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
### Load liana results by sample
# sce <- readRDS(file = "sce_with_factors.RDS")
# sce@metadata$liana_res <- readRDS(url("https://zenodo.org/record/6826420/files/pbmcs_lrs.RDS?download=1"))
```


## Cell-cell Communication Tensor Decomposition
Here, we call tensor_cell2cell.

This function will first format the ligand-receptor scores per sample into a 
4 Dimensional tensor.

It will then estimate the number of factors to which the tensor
will be decomposed (set rank is `NULL`, for the sake of computational speed,
I have pre-calculated the `rank` and I explicitly set it to 8 here). 

Finally, the function will finally decompose the tensor into a set of factors, 
each with four loadings vectors corresponding to the initial dimensions of the tensor.

One can think of this as a higher-order non-negative matrix factorization where
the factors can be used to reconstruct the tensor. We refer the user to the
publication of [tensor-cell2cell](https://www.nature.com/articles/s41467-022-31369-2) for further information.

Note that by default LIANA will set up a conda environment with `basilisk`
(if conda_env is NULL), the user can alternatively specify the name of the conda_env
to be called via reticulate.
```{r}
sce@metadata$tensor_res <-
  liana_tensor_c2c(context_df_dict = sce@metadata$liana_res,
                   score_col = 'LRscore',
                   # rank = 8,  # set to None to estimate! 8 was estimated!
                   how='inner'  #  defines how the tensor is built
                   )
```

`cell2cell_tensor` accepts additional parameters to further tune 
tensor decomposition, but these are out of scope for this tutorial.
Stay tuned for a set of tutorials with comprehensive application of liana with tensor!

### Results Overview

Here, we examine the behavior of the different dimensions across the factors.
When we look at the contexts (samples) loadings `Factor 5` seems to be notably
different between the groups. We also see that the loadings for the "Sender" and
"Receiver" cells have a relatively uniform distribution, suggesting that these most
cell types involved in the interactions that distinguish the conditions.

```{r, message=FALSE, warning=FALSE, fig.width=16, fig.height=12}
# First we format the factors
sce@metadata$tensor_res <- format_c2c_factors(sce@metadata$tensor_res)
factors <- sce@metadata$tensor_res

# Plot overview
plot_c2c_overview(factors)
```

### Contexts UMAP

Here, we will represent the contexts loadings in UMAP space. We again
observe a separation with regards to condition.
```{r}
plot_contexts_umap(factors)
```

### Statistical comparison of Communication Patterns
Here, we see that the communication patterns (or context loadings) identified 
statistically significant patterns before and after stimulation.

These factors thus represent differences in the ligand-receptor interactions
as well as cell types participating in cell-cell communication before
and after IFN-beta stimulation.


```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# Get all boxplots
all_facts_boxes <- plot_context_boxplot(factors$contexts,
                                        # paired=TRUE #! Is this the case for your data?
                                        )

# Combine all boxplots
require(patchwork)
wrap_plots(
    all_facts_boxes,
    ncol=4) +
    plot_layout(guides = "collect") & theme(legend.position = 'bottom') +
    theme(plot.tag = element_text(face = 'bold', size = 16)
          )

```



### Contexts Heatmap

Again, we see a clear seperation between STIM and CTRL, further suggesting the
relevance of the changes in the inferred interactions.
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}

plot_contexts_heat(factors)

```


### Cell-cell pairs with high potential of interaction

Here we use the product of the `source` and `target` loadings to generate the
heatmap of potential celltype-celltype pair relationships, which contribute to
`Factor.4`.
```{r ccpairs, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}

plot_c2c_cells(factors, factor_of_int = "Factor.4",
               name = "Loadings \nProduct")
```

   

### LR loadings Heatmap
We can see the LRs involved across contexts.

Though, perhaps since in this case Factors 4 and 5 are those associated with
the stimulation, perhaps it's best to focus on those.
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
plot_lr_heatmap(factors,
                loadings_thresh = 0.2,
                cluster_columns=FALSE)
```
  
### LR loadings Funcomics

Here, we will use [PROGENy](), together with [decoupleR](), to run an 
enrichment analysis on the LR loadings.

Since PROGENy is intended to be used with genes, we need to reformat it to match 
ligand-receptor predictions. Namely, `generate_lr_geneset` will assign a LR
to a specific pathway, only if both the ligand and receptor, as well as all
their subunits, are associated with the same pathway. And also in the case of
weighted gene sets (like progeny), all entities have the same weight sign.

```{r}
progeny <- decoupleR::get_progeny(organism = 'human', top=5000) %>%
    select(-p_value)
progeny_lr <- generate_lr_geneset(factors$interactions,
                                  resource = progeny)

progeny_lr
```


```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# interaction loadings to matrix
mat <- factors$interactions %>%
    column_to_rownames("lr") %>%
  as.matrix()

# run enrichment analysis with decoupler
# We don't consider genesets with minsize < 10
res <- decoupleR::run_ulm(mat = mat,
                          network = progeny_lr,
                          .source = "set",
                          .target = "lr",
                          minsize=10)

res %>% # sig/isnig flag
  mutate(significant = if_else(p_value <= 0.05, "signif.", "insig.")) %>%
  ggplot(aes(x=source, y=condition, shape=significant,
             colour=score, size=-log10(p_value+1e-36))) +
  geom_point() +
  scale_colour_gradient2(high = "red", low="blue") +
  scale_size_continuous(range = c(3, 12)) +
  scale_shape_manual(values=c(21, 16)) +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x="Pathway",
       y="Factor",
       colour="Activity"
       )

```

## Gini Coefficients of Factor-specific Communicating Sender and Receiver cell types

Gini coefficients range from 0 to 1, and are measures of dispersion,
typically used to measure inequality.

Here, the Gini coefficient is used to measure the imbalance of communication 
in terms of the cell types.

Gini coefficient of 1 would suggests that there is a single cell type contributing
to the communication patterns within a factor, while a value of 0 suggest perfect
equality between the cell type.

We will now check the contributions of Sender and Receiver cell types across
the factors.
```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# Get loadings for source/sender Cell types
calculate_gini(factors$senders)
```
When we focus on factor 6, we see that the both source/sender and target/receiver
cell types have almost uniform contributions


```{r, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# Get loadings for target/receiver Cell types
calculate_gini(factors$receivers)
```


## Citation

To Cite tensor_cell2cell:
Armingol, E., Baghdassarian, H.M., Martino, C., Perez-Lopez, A., Aamodt, C., Knight, R. and Lewis, N.E., 2022. Context-aware deconvolution of cell–cell communication with Tensor-cell2cell. Nature Communications, 13(1), pp.1-15.

To Cite LIANA:
```{r liana citation, message = FALSE, echo=FALSE}
citation("liana")
```

## Session information
```{r session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```

