---
title: "LIANA Customize CCC OmniPath"
author:
  - name: Daniel Dimitrov
    affiliation:
        - Saezlab, Heidelberg University
    email: daniel.dimitrov@uni-heidelberg.de
date: "04/08/2021"
output: 
  BiocStyle::html_document:
      self_contained: true
      toc: true
      toc_float: true
      toc_depth: 3
      code_folding: show
---

# Customize CCC OmniPath
This vignette showcases how to customize the OmniPath according to different
inbuilt filter options and additional information from OmniPath, as well as 
how to use `liana` with the customized resource.

#### Load Prerequisites
```{r load_prereq, message=FALSE, warning=FALSE}
require(liana)
require(tibble)
require(purrr)
require(magrittr)
require(dplyr)

liana_path <- system.file(package = "liana")
testdata <-
    readRDS(file.path(liana_path , "testdata", "input", "testdata.rds"))
```

## OmniPath as in the manuscript

We will first filter OmniPath as described in the LIANA paper:
* we retained only interactions with literature references,
* we kept interactions only where the receiver protein was plasma membrane transmembrane
or peripheral according to at least 30% of the localisation annotations
* we considered only interactions between single proteins (interactions between complexes are also available in OmniPath).
```{r liana_omni, message = FALSE, print = FALSE}
# generate_omni returns a tibble with CCC OmniPath
cust_omni <- generate_omni(loc_consensus_percentile = 30,
                           consensus_percentile = NULL,
                           transmitter_topology = c('secreted',
                                                    'plasma_membrane_transmembrane',
                                                    'plasma_membrane_peripheral'),
                           receiver_topology = c('plasma_membrane_transmembrane',
                                                 'plasma_membrane_peripheral'),
                           min_curation_effort = 1,
                           ligrecextra = FALSE,
                           remove_complexes = TRUE,
                           simplify = FALSE
                           )

liana_omni <- select_resource("OmniPath")[[1]]

# Check if the same
dplyr::all_equal(cust_omni, liana_omni)
```


## Surface/Membrane-bound Interactions only
```{r pm_omni, message = FALSE, print = FALSE}
pm_omni <- 
  generate_omni(loc_consensus_percentile = 51, # increase localisation consensus threshold
                consensus_percentile = NULL,
                # include only PM-bound proteins
                transmitter_topology = c('plasma_membrane_transmembrane',
                                         'plasma_membrane_peripheral'),
                receiver_topology = c('plasma_membrane_transmembrane',
                                      'plasma_membrane_peripheral'),
                min_curation_effort = 1,
                ligrecextra = FALSE,
                remove_complexes = FALSE, # keep complexes
                simplify = TRUE # do simplify
)


# check categories of ligands (category_intercell_source)
pm_omni$category_intercell_source %>% unique()

# and receptors (category_intercell_target)
pm_omni$category_intercell_target %>% unique()

# remove these categories
pm_omni %<>% filter(!(category_intercell_source %in% c("activating_cofactor",
                                                       "ligand_regulator",
                                                       "inhibitory_cofactor")))
# glimpse into the PM
pm_omni %>% glimpse()
```

## Run LIANA with custom CCC resource

#### RUN liana
```{r liana_wrap, message = FALSE, print = FALSE}
# Run liana with the custom resource
# liana Wrap
liana_test <- liana_wrap(seurat_object = testdata,
                         resource='custom',
                         external_resource = pm_omni,
                         cellchat.params=list(.normalize=TRUE)) %>%
  liana_aggregate()

liana_test %>% glimpse
```

## Under the hood
In reality, `generate_omni` is just a wrapper function that calls the appropriate `OmniPathR` functions
and simplifies their filtering according to liana-appropriate format.

```{r omnipathr, message = FALSE, print = FALSE}
# reproduce cust_op as from above with OmniPathR alone

# import the OmniPathR intercell network component
ligrec <- OmnipathR::import_intercell_network()

# filter out the complexes
ligrec %<>%
  filter(!(entity_type_intercell_source == "complex" |
             entity_type_intercell_target == "complex"))


# apply filtering according to curation and localisation
ligrec %<>% 
  OmnipathR::filter_intercell_network(loc_consensus_percentile = 30,
                                      consensus_percentile = NULL,
                                      transmitter_topology = c('secreted',
                                                               'plasma_membrane_transmembrane',
                                                               'plasma_membrane_peripheral'),
                                      receiver_topology = c('plasma_membrane_transmembrane',
                                                            'plasma_membrane_peripheral'),
                                      min_curation_effort = 1,
                                      ligrecextra = FALSE)

# remove duplicate LRs
ligrec %<>% 
  distinct_at(.vars = c("source_genesymbol", 
                              "target_genesymbol"),
                    .keep_all = TRUE)

all_equal(ligrec, cust_omni)

```
To make use of the full potential of [OmniPathR](https://github.com/saezlab/OmnipathR) we kindly refer the user to it.

It is worth noting that at this point, this tutorial only covers the intercellular component of OmniPathR, but future versions will aim to bridge this with its [intracellular part](https://omnipathdb.org/#about).

## Session information
```{r session_info, echo=TRUE}
options(width = 120)
sessioninfo::session_info()
```
