---
title: " LIANA Orhology Conversion"
author:
  - name: Daniel Dimitrov
    affiliation:
        - Saezlab, Heidelberg University
    email: daniel.dimitrov@uni-heidelberg.de
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
      self_contained: true
      toc: true
      toc_float: true
      toc_depth: 3
      code_folding: show
vignette: >
  %\VignetteIndexEntry{Homology Conversion with LIANA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

LIANA's OmniPath resource was largely compiled from CCC databases with a human focus,
as such LIANA works with human gene symbols by default.

Nevertheless, one could run LIANA with any other organism, provided a dictionary
is available to map human genesymbols to those of species X.

In this vignette, we provide an example how this could be achieved using the [`HomoloGene`](https://www.ncbi.nlm.nih.gov/homologene)
database via OmnipathR with LIANA's `generate_homologs` function. 


## Loading prerequisites
```{r load_prereq, message=FALSE, warning=FALSE}
library(tidyverse)
library(OmnipathR)
library(liana)
library(magrittr)

liana_path <- system.file(package = "liana")
testdata <-
    readRDS(file.path(liana_path, "testdata", "input", "testdata.rds"))

# Convert testdata to putative mouse symbols
# Note that we explicitly provide this tutorial to avoid any such conversions when working with real data
# We simply do this here to provide an example
rownames(testdata@assays$RNA@counts) <- stringr::str_to_title(rownames(testdata@assays$RNA@counts))
rownames(testdata@assays$RNA@data) <- stringr::str_to_title(rownames(testdata@assays$RNA@data))
```


## Internally available organisms

LIANA's `generate_homologs` function uses OmnipathR's `homologene_download` 
function to obtain a human to x species dictionary.

Here, we check the organisms which are currently available.

One must simply choose the `ncbi_taxid` and pass it as `target_organism` parameter
to the `generate_homologs` function below.
```{r}
show_homologene()
```


## Convert to Orthologues and Example LIANA run
```{r orthology, warning=FALSE, message=FALSE, fig.width=11, fig.height=8}
# Here, we will convert LIANA's Consensus resource to murine symbols
op_resource <- select_resource("Consensus")[[1]]

# Generate orthologous resource
ortholog_resource <- generate_homologs(op_resource = op_resource,
                                       target_organism = 10090) # mouse

# Run LIANA with the orthologous resource
liana_res <- liana_wrap(testdata,
                        resource = 'custom', # the resource has to be set to 'custom' to work with external resources
                        external_resource = ortholog_resource, # provide orthologous resource
                        method=c('sca', 'natmi') # run only with sca and natmi for comp. time
                        )

# aggregate
liana_res <- liana_res %>%
    liana_aggregate()

# Plot example
liana_res %>%
    filter(source=="B") %>%
    filter(target %in% c("NK", "CD8 T")) %>%
    slice_head(n=15) %>% # simply keep the highest 15 interactions according to aggregate rank
    liana_dotplot(source_groups = "B",
                  target_groups = c("NK", "CD8 T"))
```


Conversion is trickier than anticipated. In the sense that, there will be genes
without matching homologs, while other genes will have multiple matching homologs.
These issues are further complicated when working with complexes, as is the case
for the interactions in LIANA.

Furthermore, some ligand–receptor interactions could be species-specific.

Thus, we suggest that if possible to check literature for support of the predicted
edges when using orthology conversion.

Nevertheless, the `generate_homologs` function was designed to be very flexible,
but by default will be relatively conservative with conversion, with the idea
to return a resource with good conversion quality. Please refer to its documentation
for more information.



## Generating an External Dictionary
In order to convert a human resource to any other organism, we need a dictionary
which maps human gene symbols to their orthologs from another species, which 
might not be available via OmniPath. 

An alternative widely used database for this purpose is [BioMart](https://www.ensembl.org/biomart/martview/ef9109352b17a0704bd49683fe42ac81)
and we show how one could use it to obtain mouse orthologs, which we then convert to a dictionary format.

```{r biomart, eval=FALSE, warning=FALSE, message=FALSE}
# Obtain resource
op_resource <- select_resource("Consensus")[[1]]

# Get all unique symbols
symbols <- op_resource %>%
  decomplexify() %>%
  dplyr::select(ends_with("genesymbol")) %>%
  unlist() %>%
  as.vector() %>%
  unique()

# load biomaRt - an R interace for BioMart
if (!require("biomaRt", quietly = TRUE))
    install.packages("biomaRt")
library("biomaRt")

# query biomaRt databases
human <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mouse <- biomaRt::useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# Obtain tibble with human and murine genesymbol.
symbols_dict <- biomaRt::getLDS(attributes = c("hgnc_symbol"),
                                filters = "hgnc_symbol",
                                values = symbols,
                                mart = human,
                                martL = mouse,
                                attributesL = c("mgi_symbol")) %>%
  dplyr::rename(human_symbol = HGNC.symbol,
                murine_symbol = MGI.symbol) %>%
  as_tibble() %>%
  dplyr::select(human_symbol, murine_symbol) %>%
  deframe()

```

## An example dictionary (generated /w homologene) for murine symbols is included in liana
```{r hidden_load, warning=FALSE, message=FALSE}
symbols_dict <- readRDS(file.path(liana_path, "human_mouse_orthologues.RDS"))
head(symbols_dict)
```

One then simply passes the `symbols_dict` to the `symbols_dict` param instead of
specifying a tax id to `generate_homologs`.

Alternatively, one could also use external curated databases (e.g. murine 
CellChatDB) with LIANA.
