---
title: "Orthology Conversion with LIANA"
author:
  - name: Daniel Dimitrov
    affiliation:
        - Saezlab, Heidelberg University
    email: daniel.dimitrov@uni-heidelberg.de
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
      self_contained: true
      toc: true
      toc_float: true
      toc_depth: 3
      code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


LIANA's OmniPath resource was largely compiled from CCC databases with a human focus, as such
LIANA works with human symbols by default.

Nevertheless, one could, however, run LIANA with any other organism, provided a
dictionary to map human to species X. A widely used database for that purpose is BioMart. 
Here, we will show how one could use it to obtain mouse orthologs.

## Loading prerequisites
```{r load_prereq, message=FALSE, warning=FALSE}
require(tidyverse)
require(OmnipathR) 
require(liana)
require(magrittr)

liana_path <- system.file(package = "liana")
testdata <-
    readRDS(file.path(liana_path, "testdata", "input", "testdata.rds"))

# Convert testdata to putative mouse symbols
# Note that we explicitly provide this tutorial to avoid any such conversions when working with real data
# We simply do this here to provide an example
rownames(testdata@assays$RNA@counts) <- stringr::str_to_title(rownames(testdata@assays$RNA@counts)) 
rownames(testdata@assays$RNA@data) <- stringr::str_to_title(rownames(testdata@assays$RNA@data))
```

## Generate a Dictionary
```{r biomart, eval=FALSE, warning=FALSE, message=FALSE}
# Obtain resource
op_resource <- select_resource("Consensus")[[1]]

# Get all unique symbols
symbols <- op_resource %>%
  decomplexify() %>%
  dplyr::select(ends_with("genesymbol")) %>%
  unlist() %>%
  as.vector() %>% 
  unique()

# load biomaRt
if (!require("biomaRt", quietly = TRUE))
    install.packages("biomaRt")
require("biomaRt")

# query biomaRt databases
human <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mouse <- biomaRt::useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# Obtain tibble with human and murine genesymbol.
symbols_dict <- biomaRt::getLDS(attributes = c("hgnc_symbol"),
                                filters = "hgnc_symbol",
                                values = symbols,
                                mart = human,
                                martL = mouse,
                                attributesL = c("mgi_symbol")) %>%
  dplyr::rename(human_symbol = HGNC.symbol,
                murine_symbol = MGI.symbol) %>%
  as_tibble() %>%
  dplyr::select(human_symbol, murine_symbol) %>%
  deframe()
```

A dictionary for murine symbols is also included in liana
```{r hidden_load, warning=FALSE, message=FALSE}
symbols_dict <- readRDS(file.path(liana_path, "human_mouse_orthologues.RDS"))
head(symbols_dict)
```


## Convert to Orthologues and example run
```{r orthology, warning=FALSE, message=FALSE}
# Here, we will convert LIANA's Consensus resource to murine symbols
op_resource <- select_resource("Consensus")[[1]]

# Generate orthologous resource
ortholog_resource <- generate_orthologs(op_resource = op_resource,
                                        symbols_dict = symbols_dict)

# Run LIANA
liana_res <- liana_wrap(testdata,
                        resource = 'custom', # the resource has to be set to 'custom' to work with external resources
                        external_resource = ortholog_resource, # provide orthologous resource
                        method=c('sca', 'natmi') # run only with sca and natmi for comp. time
                        ) 

# aggregate
liana_res <- liana_res %>%
    liana_aggregate()

# Plot example
liana_res %>%
    filter(source=="B") %>%
    filter(target %in% c("NK", "CD8 T")) %>%
    slice_head(n=15) %>% # simply keep the highest 15 interactions according to aggregate rank
    liana_dotplot(source_groups = "B",
                  target_groups = c("NK", "CD8 T"))
```

Note, as some ligand–receptor pairs could be species specific,
we suggest to check the literature and validate reported edges when using orthology.

