---
title: "Colorectal Cancer CCC Prediction Comparison"
author: "Daniel Dimitrov"
date: "5/18/2021"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
invisible(require(intercell))
invisible(require(tidyverse))
```

#### Background
This markdown contains the results presented in the comparative Analysis of CCC predictions from different
method-resource combinations using publicly available [colorectal cancer data](https://www.nature.com/articles/s41588-020-0636-z).
Specifically, we focused on the interactions between tumour cells subclassified
by their resemblance of CRC consensus molecular subtypes (CMS) and immune cells from tumour samples.



#### Load Results from different Method-Resource Combinations
##### Note that the objects loaded here were already generated via the pipeline for each method
```{r load_res}

# We only define the measures from each method that we wish to have in our analysis
spec_list <- list("CellChat" =
                      methods::new("MethodSpecifics",
                                   method_name="CellChat",
                                   method_results = readRDS("output/crc_res/cellchat_results.rds"),
                                   method_scores=list(
                                       "prob"=TRUE
                                       )),
                  "Connectome" =
                      methods::new("MethodSpecifics",
                                   method_name="Connectome",
                                   method_results = readRDS("output/crc_res/conn_results.rds"),
                                   method_scores=list(
                                       "weight_sc"=TRUE
                                   )),
                  "iTALK" =
                      methods::new("MethodSpecifics",
                                   method_name="iTALK",
                                   method_results = readRDS("output/crc_res/italk_results.rds"),
                                   method_scores=list(
                                       "weight_comb"=TRUE
                                   )),
                  "NATMI" =
                      methods::new("MethodSpecifics",
                                   method_name="NATMI",
                                   method_results = readRDS("output/crc_res/natmi_results.rds"),
                                   method_scores=list(
                                       "edge_specificity"=TRUE
                                       )),
                  "SCA" = methods::new("MethodSpecifics",
                                       method_name="SCA",
                                       method_results = readRDS("output/crc_res/sca_results.rds"),
                                       method_scores=list(
                                           "LRscore"=TRUE
                                           )),
                  "Squidpy" =
                      methods::new("MethodSpecifics",
                                   method_name="Squidpy",
                                   method_results = readRDS("output/crc_res/squidpy_results.rds"),
                                   method_scores=list(
                                       "pvalue"=FALSE
                                   ))
                  )
# Define the numbers of highest interactions that we wish to explore
# and get a list with each threshold as its element
top_lists <- get_top_hits(spec_list,
                          n_ints=c(100,
                                   250,
                                   500,
                                   1000)
                          )
```


### Main Text Plots
#### Combine all binary results into heatmap (top500)
Overlap in the 500 highest ranked CCC interactions between different combinations
of methods and resources. Method-resource combinations were clustered according 
to binary (Jaccard index) distances. SCA refers to the SingleCellSignalR method.
```{r binary_heat_main, fig.width=46, fig.height=23}

png(filename = figure_path_mr('crc_binheat_top500.png'),
    width = 3600,
    height = 1700)

p <- get_BinaryHeat(top_lists$top_500)
print(p)
invisible(dev.off())
grid::grid.draw(p$gtable)
```

#### Activity per Cell type
##### Inferred as the proportion of interaction edges that stem from Source Cell clusters or lead to Target Cell clusters in the highest ranked interactions. 
```{r binary_cell_activity_main, fig.width=46, fig.height=23}

png(filename = figure_path_mr('crc_activityheat_top500.png'),
    width = 3600,
    height = 1700)

# Cap to 0.2 (due to high activity in SCA)
p <- get_activecell(top_lists$top_500, cap_value = 0.2)
invisible(dev.off())
grid::grid.draw(p$gtable)
```

#### Jaccard Index exploration
```{r jacc_index, fig.width=46, fig.height=23}
# Similarity Heatmap (according to Jaccard index)
p <- get_simdist_heatmap(top_lists$top_500,
                         sim_dist = "simil",
                         method = "Jaccard",
                         diag = TRUE,
                         upper = TRUE)

png(filename = figure_path_mr('crc_jaccard_top500.png'),
    width = 3600,
    height = 1700)
print(p)
invisible(dev.off())
print(p)

# Get Jaccard Stats
jaccard_per_mr <- simdist_resmet(top_lists$top_500,
                                 sim_dist = "simil",
                                 method = "Jaccard")
jac <- list_stats(meth = jaccard_per_mr$meth,
                  reso = jaccard_per_mr$reso)

# Here we obtain the mean Jaccard index per resource when using different methods
# and per method when using different resources
jac


# pairwise JI between methods
methods <- c("CellChat", "Connectome", "iTALK", "NATMI", "SCA", "Squidpy")
methods_jacc <- methods %>%
    combn(2) %>%
    t() %>%
    as_tibble() %>%
    unite(c("V1", "V2"), col = "method_combo") %>%
    mutate(methods = str_split(method_combo, "_")) %>%
    mutate(jacc_mat = pmap(list(.x = method_combo,
                                .y = methods),
                           .f = function(.x, .y){
                               list(.x = get_jacc(top_lists$top_500,
                                                  .y,
                                                  as.character(get_lr_resources()))
                                    )
                                    }
                           )
           ) %>%
    unnest(jacc_mat) %>%
    rowwise() %>%
    mutate(jacc_mean = mean(jacc_mat)) %>%
    ungroup() %>%
    arrange(desc(jacc_mean))

# pairwise similarity between each method combination
# i.e. here we don't consider the universe of method-resource combinations
# but rather just the comparison of any two methods when using each resource
methods_jacc


# Jaccard Index using the Same Method with different resources
#  (i.e. similarity among combinations using the same method)
# or in other words how similar are the results from each method with different resources
methods %>% map(function(met){
    get_jacc(top_lists$top_500,
             met,
             as.character(as.character(get_lr_resources()))) %>%
        mean()
}) %>% setNames(methods)


```


#### Check hits per method (whenever a threshold is proposed)
```{r check_numbers}
# Check CellChat P-values
cc_hits <- spec_list$CellChat@method_results %>%
    map(function(resource) resource %>%
            filter(pval <= 0.05))
cc_hits %>% map(function(x) nrow(x)) %>% enframe(name="method", value="n_hits") %>% unnest(n_hits)

# Check SCA above threshold
sca_hits <- spec_list$SCA@method_results %>%
    map(function(resource) resource %>%
            filter(LRscore >= 0.5))
sca_hits %>% map(function(x) nrow(x)) %>% enframe(name="method", value="n_hits") %>% unnest(n_hits)

# Check Squidpy P-values
sq_hits <- spec_list$Squidpy@method_results %>%
    map(function(resource) resource %>%
            filter(pvalue <= 0.05))
sq_hits %>% map(function(x) nrow(x)) %>% enframe(name="method", value="n_hits") %>% unnest(n_hits)
```



### Supplementary Plots
####  Upset Plots by Method for highest ranked 500 interactions (CRC)
```{r upsets_by_method, fig.width=36, fig.height=18}
names(top_lists$top_500) %>%
    map(function(m_name){
        top_lists$top_500[[m_name]] %>%
            prepForUpset() %>%
            plotSaveUset(figure_path_mr(as.character(str_glue("crc_{m_name}_top500_upset_method.png"))),
                         m_name)
        }
    )
```

####  Upset Plots by Resource for highest ranked 500 interactions (CRC)
```{r upsets_by_resource, fig.width=36, fig.height=18}
top250_resource_tool <- get_swapped_list(top_lists$top_500)

# Plot and Save Upsets
names(top250_resource_tool) %>%
    map(function(r_name){
        top250_resource_tool[[r_name]] %>%
            prepForUpset() %>%
            plotSaveUset(figure_path_mr(as.character(str_glue("crc_{r_name}_top500_upset_resource.png"))),
                         r_name)
        }
    )

```

#### Binary Jaccard index heatmaps (Supp interactions: 100, 250, and 1000)
```{r binary_heat_supp, fig.width=32, fig.height=17}

png(filename = figure_path_mr('crc_binheat_top100.png'),
    width = 3600,
    height = 1660)
get_BinaryHeat(top_lists$top_100)
invisible(dev.off())

png(filename = figure_path_mr('crc_binheat_top250.png'),
    width = 3600,
    height = 1600)
get_BinaryHeat(top_lists$top_250)
invisible(dev.off())

png(filename = figure_path_mr('crc_binheat_top1000.png'),
    width = 3600,
    height = 1700)
get_BinaryHeat(top_lists$top_1000)
invisible(dev.off())

```


#### Number of Cell Types in the CRC dataset after filtering
```{r cellnums}
# Load Seurat and Check numbers from metadata
get_cellnum("input/crc_data/crc_korean_form.rds")
```


#### PCA by Interaction Rank Frequencies
i.e. Interaction ranks per Pairs of Cell types z-normalized
```{r pca_rank_activities, fig.width=24, fig.height=12}
p <- spec_list %>%
    get_rank_frequencies() %>%
    plot_freq_pca()

png(filename = figure_path_mr('crc_rank_activity_pca.png'),
    width = 1920,
    height = 1024)

print(p)
invisible(dev.off())

p

```


### Supplementary Note 2
#### Conversion Differences
```{r conversion_diff, fig.width=32, fig.height=17}

# Jaccard index between the Inbuilt resource for each method
# and the omnipath querried counterpart of the same resource
methods_default <- list("iTALK" = "iTALK",
                        "CellChat" = "CellChatDB",
                        "Connectome" = "Ramilowski2015",
                        "NATMI" = "connectomeDB2020",
                        "SCA" = "LRdb")

methods_default %>%
    enframe(name = "method", value = "default") %>%
    unnest(default) %>%
    mutate(default = as.character(str_glue("{method}_{default}"))) %>%
    mutate(inbuilt = stringr::str_replace(default, "\\_..*", "_Default")) %>%
    mutate(jacc = pmap_dbl(list(default, inbuilt),
                       .f = function(x, y){
                           get_binary_df(top_lists$top_1000) %>%
                               select(c(rlang::as_string(x), rlang::as_string(y))) %>%
                               t() %>%
                               get_simil_dist(sim_dist = "simil", "Jaccard")
                           }
                       )
           )




```

#### Protein Complexes


### Supplementary Note 3 
#### Cluster Specificity and Method Dissimilarity
```{r nonspec_load}
# Load method results with non-specific cluster measures
nonspec_list <- list("CellChat" =
                         methods::new("MethodSpecifics",
                                   method_name="CellChat",
                                   method_results = readRDS("output/crc_res/cellchat_results.rds"),
                                   method_scores=list(
                                       "prob"=TRUE
                                       )),
                      "Connectome" =
                      methods::new("MethodSpecifics",
                                   method_name="Connectome",
                                   method_results = readRDS("output/crc_res/conn_results.rds"),
                                   method_scores=list(
                                       "weight_norm"=TRUE
                                   )),
                      "NATMI" =
                      methods::new("MethodSpecifics",
                                   method_name="NATMI",
                                   method_results = readRDS("output/crc_res/natmi_results.rds"),
                                   method_scores=list(
                                       "edge_avg_expr"=TRUE
                                   )),
                      "Squidpy" =
                      methods::new("MethodSpecifics",
                                   method_name="Squidpy",
                                   method_results = readRDS("output/crc_res/squidpy_results.rds"),
                                   method_scores=list(
                                       "means"= TRUE
                                   ))
)

nonspec_top_list <- get_top_hits(nonspec_list,
                              n_ints=c(500))

```


#### Produce plots
```{r nonspec_res, fig.width=32, fig.height=17}
# Interactions Overlap Binary Heatmap
png(filename = figure_path_mr('crc_nonspec_binheat_top500.png'),
    width = 3600,
    height = 1700)

p <- get_BinaryHeat(nonspec_top_list$top_500)
grid::grid.draw(p$gtable)
invisible(dev.off())
p


# Activity per Cell Type Heatmap
png(filename = figure_path_mr('crc_nonspec_activityheat_top500.png'),
    width = 3600,
    height = 1700)

# Cap to 0.2
p <- get_activecell(nonspec_top_list$top_500)
grid::grid.draw(p$gtable)
invisible(dev.off())
grid::grid.draw(p)

# Jaccard Stats Heatmap
get_simdist_heatmap(nonspec_top_list$top_500,
                    sim_dist = "simil",
                    method = "Jaccard",
                    diag = TRUE,
                    upper = TRUE)

jaccard_nonspec <- simdist_resmet(nonspec_top_list$top_500,
                                 sim_dist = "simil",
                                 method = "Jaccard")
# Mean Jacc Index per Method and Resource
list_stats(meth = jaccard_nonspec$meth,
           reso = jaccard_nonspec$reso)

```
