---
title: "panc8_analysis"
author: "Daniel Dimitrov"
date: "5/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(intercell)
require(tidyverse)
```

### Background
scRNA-Seq data was aquired via SeuratData for human [pancreatic cells](https://www.nature.com/articles/nbt.4096)
   
   
#### Load Results from different Method-Resource Combinations
##### Note that the objects loaded here were already generated via the pipeline for each method
```{r load_res}

# We only define the measures from each method that we wish to have in our analysis
spec_list <- list("CellChat" =
                      methods::new("MethodSpecifics",
                                   method_name="CellChat",
                                   method_results = readRDS("output/panc8_res/cellchat_results.rds"),
                                   method_scores=list(
                                       "prob"=TRUE
                                       )),
                  "Connectome" =
                      methods::new("MethodSpecifics",
                                   method_name="Connectome",
                                   method_results = readRDS("output/panc8_res/conn_results.rds"),
                                   method_scores=list(
                                       "weight_sc"=TRUE
                                   )),
                  "iTALK" =
                      methods::new("MethodSpecifics",
                                   method_name="iTALK",
                                   method_results = readRDS("output/panc8_res/italk_results.rds"),
                                   method_scores=list(
                                       "weight_comb"=TRUE
                                   )),
                  "NATMI" =
                      methods::new("MethodSpecifics",
                                   method_name="NATMI",
                                   method_results = readRDS("output/panc8_res/natmi_results.rds"),
                                   method_scores=list(
                                       "edge_specificity"=TRUE
                                       )),
                  "SCA" = methods::new("MethodSpecifics",
                                       method_name="SCA",
                                       method_results = readRDS("output/panc8_res/sca_results.rds"),
                                       method_scores=list(
                                           "LRscore"=TRUE
                                           )),
                  "Squidpy" =
                      methods::new("MethodSpecifics",
                                   method_name="Squidpy",
                                   method_results = readRDS("output/panc8_res/squidpy_results.rds"),
                                   method_scores=list(
                                       "pvalue"=FALSE
                                   ))
                  )
# Define the numbers of highest interactions that we wish to explore
# and get a list with each threshold as its element
top_lists <- get_top_hits(spec_list,
                          n_ints=c(100,
                                   250,
                                   500,
                                   1000)
                          )
```


### Main Text Plots
#### Combine all binary results into heatmap (top500)
Overlap in the 500 highest ranked CCC interactions between different combinations
of methods and resources. Method-resource combinations were clustered according 
to binary (Jaccard index) distances. SCA refers to the SingleCellSignalR method.
```{r binary_heat_main, fig.width=46, fig.height=23}

png(filename = figure_path_mr('panc_binheat_top500.png'),
    width = 3000,
    height = 1680)

p <- get_BinaryHeat(top_lists$top_500)
grid::grid.draw(p$gtable)
invisible(dev.off())
grid::grid.draw(p$gtable)
```
   
   
  
#### Activity per Cell type
##### Inferred as the proportion of interaction edges that stem from Source Cell clusters or lead to Target Cell clusters in the highest ranked interactions. 
```{r binary_cell_activity_main, fig.width=46, fig.height=23}
p <- get_activecell(top_lists$top_500)

png(filename = figure_path_mr('panc_activityheat_top500.png'),
    width = 3000,
    height = 1700)

grid::grid.draw(p$gtable)
invisible(dev.off())

grid::grid.draw(p$gtable)
```
   
   
   
#### Jaccard index Heatmap
```{r jacc_index, fig.width=46, fig.height=23}
# Similarity Heatmap (according to Jaccard index)
p <- get_simdist_heatmap(top_lists$top_500,
                         sim_dist = "simil",
                         method = "Jaccard",
                         diag = TRUE,
                         upper = TRUE)

png(filename = figure_path_mr('panc_jaccard_top500.png'),
    width = 3200,
    height = 2800)
grid::grid.draw(p$gtable)
invisible(dev.off())
grid::grid.draw(p$gtable)
```
